\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2009-07-14 papers/jss/article.tex
added EM bit
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@297 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
some fixes for the jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@298 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% an abstract and keywords
\Abstract{ 	
	depmixS4 implements a general framework for definining and
	estimating hidden Markov mixture model in the R programming language
	R2009.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach using the Rdonlp2 optimization
	routine when contraints are imposed on the parameters.  A number
	of illustrative examples are included.
}

\Keywords{hidden markov model, dependent mixture model, mixture model}

@@

\newcommand{\vc}{\mathbf}
\newcommand{\mat}{\mathbf}

@@

Need some more on EM and how/why it is justified to do separate weighted
iuuuuuuuuuuucvfits of the response models and transition and prior models.

Also mention use of glm, nnet and possibly other packages that we use.

\section{Using \pkg{depmixS4}}

@@

\begin{verbatim} 
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
\end{verbatim}
The \code{depmix} function returns an object of class \code{depmix}
which contains the model specification (and not a fitted model!).
Note also that start values for the transition parameters are provided
in this call using the \code{trstart} argument. The package does not 
provide automatic starting values.

The so-defined models needs to be \code{fit}ted with the following: 
\begin{verbatim}
fm <- fit(mod)
\end{verbatim}

@@

\begin{verbatim}
> fm 
Convergence info: Log likelihood converged to within tol. 
'log Lik.' -84.34175 (df=7)
AIC:  182.6835 
BIC:  211.275 
\end{verbatim}

@@

\begin{verbatim}
> summary(fm)
Initial state probabilties model 
Model of type multinomial, formula: ~1
Coefficients: 
     [,1]      [,2]
[1,]    0 -11.25688
Probalities at zero values of the covariates.
0.999987 1.291797e-05

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 5.511151
sd  0.1926063 
\end{verbatim}

@@

Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{verbatim}
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, 
family=gaussian(), transition=~scale(Pacc), 
instart=runif(2))
fm <- fit(mod)
\end{verbatim}

\begin{verbatim}
	Initial state probabilties model 
	Model of type multinomial, formula: ~1
	Coefficients: 
		 [,1]     [,2]
	[1,]    0 10.71779
	Probalities at zero values of the covariates.
	2.214681e-05 0.9999779

@@

	Response model for response 1 
	Model of type gaussian, formula: rt ~ 1
	Coefficients: 
	[1] 6.3885
	sd  0.2402693

\end{verbatim}

@@

\section{Extending \pkg{depmixS4}}

@@

The \code{fit} function is defined as follows: 
\begin{verbatim}
setMethod("fit","exgaus",
	function(object,w) {
		if(missing(w)) w <- NULL
		y <- object@y
		fit <- gamlss(y~1,weights=w,family=exGAUS(),
			control=gamlss.control(n.cyc=100,trace=FALSE),
			mu.start=object@parameters$mu,
			sigma.start=exp(object@parameters$sigma),
			nu.start=exp(object@parameters$nu))
		pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
		object <- setpars(object,pars)
		object
	}
)
The \code{fit} function defines a trivial \code{gamlss} model with 
only an intercept to be estimated and then sets the fitted parameters 
back into their respective slots in the `exgaus' object. See the help 
for \pkg{gamlss.distr} for interpretation of these parameters.

@@

\section{Conclusion \& future work}

>>>>

%% an abstract and keywords
\Abstract{ 	
	\pkg{depmixS4} implements a general framework for definining and
	estimating hidden Markov mixture model in the R programming language
	\citep{R2009}.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach using the \pkg{Rdonlp2} optimization
	routine when contraints are imposed on the parameters.  A number
	of illustrative examples are included.
}

\Keywords{hidden Markov model, dependent mixture model, mixture model}

@@

\newcommand{\vc}{\mathbf}
\newcommand{\mat}{\mathbf}
\newcommand{\greekv}[1]{\mbox{\boldmath$\mathrm{#1}$}}
\newcommand{\greekm}[1]{\mbox{\boldmath$#1$}}

@@

%Need some more on EM and how/why it is justified to do separate weighted
%iuuuuuuuuuuucvfits of the response models and transition and prior models.

%Also mention use of glm, nnet and possibly other packages that we use.

\section[Using depmixS4]{Using \pkg{depmixS4}}

@@

\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
\end{CodeInput}
\end{CodeChunk}

The \code{depmix} function returns an object of class \code{depmix}
which contains the model specification (and not a fitted model!).
Note also that start values for the transition parameters are provided
in this call using the \code{trstart} argument. The package does not 
provide automatic starting values.

The so-defined models needs to be \code{fit}ted with the following: 
\begin{CodeChunk}
\begin{CodeInput}
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
> fm
\end{CodeInput}
\begin{CodeOutput}
Convergence info: Log likelihood converged to within tol. 
'log Lik.' -84.34175 (df=7)
AIC:  182.6835 
BIC:  211.275 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
Initial state probabilties model 
Model of type multinomial, formula: ~1
Coefficients: 
     [,1]      [,2]
[1,]    0 -11.25688
Probalities at zero values of the covariates.
0.999987 1.291798e-05

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 5.511151
sd  0.1926063 
\end{CodeOutput}
\end{CodeChunk}

@@

Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
  transition=~scale(Pacc), instart=runif(2))
> fm <- fit(mod)
\end{CodeInput}
\begin{CodeOutput}
Initial state probabilties model 
Model of type multinomial, formula: ~1
Coefficients: 
	 [,1]     [,2]
[1,]    0 10.71779
Probalities at zero values of the covariates.
2.214681e-05 0.9999779

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 6.3885
sd  0.2402693 
\end{CodeOutput}
\end{CodeChunk}

@@

\section[Extending depmixS4]{Extending \pkg{depmixS4}}

@@

The \code{fit} function is defined as follows: 
\begin{verbatim}
setMethod("fit","exgaus",
	function(object,w) {
		if(missing(w)) w <- NULL
		y <- object@y
		fit <- gamlss(y~1,weights=w,family=exGAUS(),
			control=gamlss.control(n.cyc=100,trace=FALSE),
			mu.start=object@parameters$mu,
			sigma.start=exp(object@parameters$sigma),
			nu.start=exp(object@parameters$nu))
		pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
		object <- setpars(object,pars)
		object
	}
)
\end{verbatim}

The \code{fit} function defines a trivial \code{gamlss} model with 
only an intercept to be estimated and then sets the fitted parameters 
back into their respective slots in the `exgaus' object. See the help 
for \code{gamlss.distr} for interpretation of these parameters.

@@

\section[Conclusion and future work]{Conclusion \& future work}
